\documentclass[%
%prl%  
pra%
%rmp%
% ,preprint%
,twocolumn%
%,secnumarabic%
%,tightenlines%
%,byrevtex 
,amssymb, nobibnotes, aps,
longbibliography
]{revtex4-1}
%
\usepackage{amsmath}
\usepackage{euscript}
\usepackage{graphicx}
\usepackage[usenames]{color}
\usepackage{bbold}
%
\begin{document}
\title{
Double photoionization of a triatomic molecule  treating two active electrons: DVR and orbital basis
}
\author{CWM begun June 2020 }
%
\begin{abstract}
These notes will lay out all aspects of double photoionization calculations on water
\begin{itemize}
\item {Basic formalism}
\item {Representation of $\Psi_{sc}^+$ and $\Phi_0$}
\item {transforming from DVR to orbital basis}
\item{ extracting double ionization amplitudes}
\end{itemize}   
\end{abstract}
%
\maketitle
\tableofcontents
%
%
%
\section{Formulation of the problem in a single center DVR representation}
\label{sec:formulation}

We start with a reduction of the full double ionization problem to an effective two electron problem, treating two active electrons.  For example suppose we are treating one-photon double ionization to produce  the 1$^1$A$_1$ of H$_2$O$^++$.  If we freeze all the orbitals but the $1b_1$ orbital in the ground state configuration, $1a_1^2 \,  2a_1^2 \, 1b_2^2 \, 3a_1^2 \, 1b_1^2$ we would define the initial state including only correlation between the two $1b_1$ electrons as the singles and doubles CI wave function taking excitations only out of the two $1b_1$ spin orbitals.  If we express that wave function in terms of the natural orbitals of the singles and doubles CI, a theorem due to L\"owdin states that it has the form 
\begin{equation}
\begin{split}
\Phi_{gnd}   =& \sum_{n} C_{n}  |1a_1^2 \,  2a_1^2 \, 1b_2^2 \, 3a_1^2 \, n\alpha \, n\beta  |
\end{split}
\end{equation}
where there are only doubly occuppied natural orbitals in the list of determinants.  So we can pick the initial two electron state for our two-active-electrons description of the double ionization process as
\begin{equation}
\begin{split}
\Phi_0(1,2) &=  \sum_{n} C_{n}  |n\alpha(1)  \, n\beta(2)  | \\
& =  \sum_{n} C_{n} \phi_n(\mathbf{r}_1) \phi_n(\mathbf{r}_2) \frac{\alpha(1)\beta(2)-\beta(1)\alpha(2))}{\sqrt{2}}
\end{split}
\end{equation}
which we can represent on the DVR grid using the single center expansion of the natural orbitals
\begin{equation}
\phi_n(\mathbf{r}) = \sum_{l,m} R^{(n)}_{lm}(r) X_{lm} (  \mathbf{\hat{r}} )
\end{equation}
where $X_{lm}$ denotes a real spherical harmonic
\begin{equation}
\begin{split}
\Phi_0(1,2) &=  \sum_{n} C_{n}  |n\alpha(1)  \, n\beta(2)  | \\
& =  \sum_{n} C_{n} \phi_n(\mathbf{r}_1) \phi_n(\mathbf{r}_2) \\
& =  \sum_{\ell,m,\ell',m'} \left[  \sum_{n} C_{n} R^{(n)}_{\ell m}(r_1)  R^{(n)}_{\ell' m'}(r_2)\right]X_{\ell m} (  \mathbf{\hat{r}}_1 )X_{\ell' m'} (  \mathbf{\hat{r}}_2 ) \\
 \textrm{or} \\
\Phi_0(\mathbf{r}_1,\mathbf{r}_2) &= \sum_{\ell,m,\ell',m'} \EuScript{R}_{\ell m; \ell' m'}(r_1,r_2) \, X_{\ell m} (  \mathbf{\hat{r}}_1 )X_{\ell' m'} (  \mathbf{\hat{r}}_2 ) 
\end{split}
\end{equation}
where the radial functions are defined by the expression in square brackets, and clearly have the symmetries $\EuScript{R}_{\ell m ;\ell' m'}(r_1,r_2)=\EuScript{R}_{\ell m; \ell' m'}(r_2,r_1)$ and $\EuScript{R}_{\ell m; \ell' m'}(r_1,r_2)=\EuScript{R}_{\ell' m' ;\ell m}(r_1,r_2)$.
\\
\


\noindent \textbf{Code Check:}  in the DVR basis the two-electron radial functions are
\begin{equation}
\begin{split}
 \EuScript{R}_{\ell,m;\ell',m'} (r_1,r_2) = \sum_{i;j}  \EuScript{R}_{\ell,m;\ell',m'}^{i;j} \frac{\chi_i(r_1)}{r_1} \frac{ \chi_j(r_2) }{r_2}
\end{split}
\end{equation}
So the one-electron density matrix in this representation is
\begin{equation}
\begin{split}
\rho_{i,\ell,m;i',\ell',m'} = \sum_{j',\ell'',m''} \EuScript{R}_{\ell,m;\ell'',m''}^{i;j'}   \EuScript{R}_{\ell',m';\ell'',m''}^{i';j'} 
\end{split}
\end{equation}
So after constructing the representation of $\Phi_0$ from Gaussian orbitals, we can construct the density matrix and diagonalize it to get back the same naural orbital occupations as its eigenvalues that we started with from the Gaussian natural orbitals.  If there are N natural orbitals, this large matrix should have only N nonzero eigenvalues, to within the various numerical errors of the re-expansion of the Gaussian orbitals in the DVR basis.
\\
\


With the right hand side constructed with this function we solve the driven Schr\"odinger equation for the scattered wave
\begin{equation}
(E_0+\hbar \omega -H_{eff})\Psi_{sc}^+(\mathbf{r}_1,\mathbf{r}_2) = \mu \Phi_0(\mathbf{r}_1,\mathbf{r}_2)
\label{eq:drivenSE}
\end{equation}
with, in this case,
\begin{equation}
H_{eff} = \hat{T}_1 + \hat{T}_2 + V_{nuc} + \sum_o \left( 2 \hat{J}_o - \hat{K}_o \right) + \frac{1}{r_{12}} 
\end{equation}
and then extract the double ionization amplitudes from $\Psi_{sc}^+(\mathbf{r}_1,\mathbf{r}_2) $, which may be problematic. 
We can also time propagate to obtain 
\begin{equation}
\Psi(\mathbf{r}_1,\mathbf{r}_2,t) = \exp[-iH_{eff})t]\,\mu \Phi_0(\mathbf{r}_1,\mathbf{r}_2)
\end{equation}
 and project on it with a product of Coulomb functions at long enough times to extract the double ionization amplitudes.                                                                                                                  

\section{Orbital basis in place of the DVR}
\label{sec:orbbasis}

\subsection{Nonspherical orbital basis: Method 1 requiring diagonalization of full one-body FEM-DVR Hamiltonian}
\label{sec:nonspherical}

We would need to solve Eq.(\ref{eq:drivenSE}) in a space orthogonal to the doubly occupied orbitals that were frozen in this model.  We could do that by constructing the DVR representation of a projection on the doubly occupied orbitals
\begin{equation}
\begin{split}
\hat{P}_o(i) & \equiv  \langle \mathbf{r}_i | \phi_o \rangle \langle \phi_o|\mathbf{r'}_i \rangle \\
\hat{Q}_o &= 1 - \hat{P}_o(1) -  \hat{P}_o(2 ) + \hat{P}_o(1)\hat{P}_o(2)
\end{split}
\end{equation}
The $\hat{Q}_o$ operator is a two-electron projector that projects out orbital $\phi_o$. The full projector to project out a set of orbitals would then be
\begin{equation}
\begin{split}
\hat{Q}&= 1 - \sum_o (  \hat{P}_o(1) +  \hat{P}_o(2 ) - \hat{P}_o(1)\hat{P}_o(2))
\end{split}
\end{equation}
Which we would build in the DVR, again using $\phi_o(\mathbf{r}) = \sum_{l,m} R^{(o)}_{lm}(r) X_{lm} (  \mathbf{\hat{r}} )$.  To use it we would have to operate with $Q$ to construct $\hat{Q} H_{eff} \hat{Q}$ and operate in that rank defficient DVR representation.  This seems to be what Frank Yip did in the atomic cases he has done so far.

Instead of doing this, since we may want to project out not only the doubly occupied orbitals but also parts of the DVR space that corresponds to very high kinetic energy or one-body eigenvalues, suppose we construct a complete orbital basis and then represent Eq.(\ref{eq:drivenSE}) in that orbital space?  What would that entail?

First we have to construct a complete orbital basis, consisting of the doubly occupied orbitals and the eigenfunctions of the one-body Hamiltonian in that space.
The one-body operator is
\begin{equation}
h(\mathbf{r})= \hat{T}_1 + V_{nuc}(\mathbf{r})  + \sum_o \left( 2 \hat{J}_o(\mathbf{r}) - \hat{K}_o(\mathbf{r})  \right) 
\end{equation}
Their are M doubly occupied orbitals, so to project them out of the space we need
\begin{equation}
\begin{split}
\phi_o(\mathbf{r}) &= \sum_{\ell,m} R^{(o)}_{\ell m}(r) X_{\ell m} (  \mathbf{\hat{r}} )\\
\hat{Q}&= 1 - \sum_o \sum_{\ell,m} R^{(o)}_{\ell m}(r) X_{\ell m} (  \mathbf{\hat{r}} ) \sum_{\ell',m'} R^{(o)}_{\ell' m'}(r') X_{\ell' m'} (  \mathbf{\hat{r'}} ) \\
&= 1- \sum_{\ell,m,\ell',m'} \left[\sum_{o=1}^M  R^{(o)}_{\ell m}(r) R^{(o)}_{\ell' m'}(r') \right] X_{\ell m} (  \mathbf{\hat{r}} ) X_{\ell' m'} (  \mathbf{\hat{r'}} ) \\
&= 1- \sum_{\ell,m,\ell',m'}  \EuScript{R}_{\ell,m,\ell',m'} (r,r') X_{\ell m} (  \mathbf{\hat{r}} ) X_{\ell' m'} (  \mathbf{\hat{r'}} )
\end{split}
\label{eq:Qonebody}
\end{equation}
In the radial DVR basis, which we will assume here is the same for all $\ell, m$, the two-electron radial functions that define the projector have the representation
\begin{equation}
\begin{split}
 \EuScript{R}_{\ell,m,\ell',m'} (r,r') = \sum_{i,j}  \EuScript{R}_{\ell,m,\ell',m'}^{i,j} \frac{\chi_i(r)}{r} \frac{ \chi_j(r') }{r'}
\end{split}
\end{equation}

So in the single center DVR basis we are using, $\chi_i(r) X_{\ell,m} (\mathbf{\hat{r}})$ The Q operator is the matrix
\begin{equation}
\begin{split}
\hat{Q}_{i,\ell,m ;  j, \ell',m'} 
&= \delta_{\ell,\ell'} \delta_{m,m'} \delta_{i,j}  - \EuScript{R}_{\ell,m,\ell',m'}^{i,j} 
\end{split}
\end{equation}
If we have the one-body Hamiltonian in this primitive DVR basis, then transformed Hamiltonian
\begin{equation}
\begin{split}
& \tilde{h}_{i,\ell,m, j, \ell',m'} = \\
& \sum_{ k, \ell'',m'', l,\ell''',m'''} \hat{Q}_{i,\ell,m, k, \ell'',m''}\, h_{k,\ell'',m'', l, \ell''',m'''} \,\hat{Q}_{l,\ell''',m''', j, \ell',m'} 
 \end{split}
\end{equation}
should have exactly $M$ zero eigenvalues, and the rest of the eigenvectors, 
\begin{equation}
\begin{split}
& \sum_{j,\ell',m'}\tilde{h}_{i,\ell,m, j, \ell',m'} R^{\alpha}_{j,\ell'.m'} = \epsilon_\alpha \,R^{\alpha}_{i,\ell.m} \\
&\phi_\alpha(\mathbf{r}) = \sum_{i,\ell,m} \,R^{\alpha}_{i,\ell.m}\,  \frac{\chi_i(r)}{r} X_{\ell,m}
 \end{split}
\end{equation}
augmented by the occupied orbitals, would be the complete orbital basis.
\\
\

\noindent \textbf{Code check:} There are 4 doubly occupied orbitals in the water case, but to check project out just 1, 2, 3 and 4 of them to check that the number of zero eigenvalues is the number of orbitals projected out.
\\
\

This is a completely equivalent basis to our original DVR basis and the orbitals $\phi_\alpha(\mathbf{r})$ are nonspherical.  But they also don't have any particular physical meaning (except for the occupied orbitals).  Moreover the size of the matrices involved is large because all the $\ell,m$ pairs are coupled in principle.  Also the one-body Hamiltonian is a large matrix, and we are requiring all of its eigenvectors, which may be problematic.  Since we are not using any physical properties of these functions except their eigenvalues, can we make orbital basis functions by diagonalizing the one-body Hamiltonian  for each $\ell,m$ pair separately?

\subsection{Nonspherical orbital basis: Method 2 using eigenvectors of each $\ell,m$ block of the one-body Hamiltonian}
\label{sec:lorbs}



\subsubsection{Kinetic energy:}  
The real spherical harmonics are eigenfunctions of $\hat{L}^2$ so the kinetic energy matrix
elements don't change from their form for complex spherical harmonics.
\begin{equation}
T_{i,\ell,m;j,\ell',m'}^{(\ell)} =  T_{i,j}^{(\ell)}  \delta_{\ell,\ell'} \delta_{m,m'}
\end{equation}
with
\begin{equation}
T_{n,m}^{(\ell)} = -\frac{1}{2} \int_0^{r_{max}} \chi_n(r)
\left(\frac{d^2}{dr^2}-\frac{\ell(\ell+1)}{r^2}\right)
\chi_m(r) dr
\end{equation}
\\
\

\subsubsection{Nuclear attraction potential}
The nuclear attraction potential is
\begin{equation}
V_{nuc}(\mathbf{r}) = \sum_\beta \frac{Z_\beta}{|\mathbf{r}-\mathbf{R}_\beta|}
\end{equation}
To perform the integrals 
\begin{equation}
\begin{split}
& V^{nuc}_{i, \, \ell, \,m; \, j, \,\ell', \, m'}(\mathbf{R}_\beta) = \\
&\int \frac{\chi_{i}(r)}{r}X_{\ell, \,m} (\mathbf{\hat{r}})
\frac{1}{|\mathbf{r}-\mathbf{R}_\beta|}
\frac{\chi_j(r)}{r} X_{\ell', \, m'} (\mathbf{\hat{r}})
d\mathbf{r}
\end{split}
\end{equation}
we need the multipole expansion in terms of the real spherical harmonics, which is
\begin{equation}
\frac{1}{| \mathbf{r}- \mathbf{r}'|} = 
 \sum_{\lambda,\mu} \frac{4 \pi}{2\lambda +1}
X_{\lambda,\mu}(\mathbf{\hat{r}})
X_{\lambda,\mu}(\mathbf{\hat{r}}')
\frac{r^\lambda}{r^{\lambda+1}}
\end{equation}
as shown in Appendix \ref{sec:angmom}.

We also need the integral of three real spherical harmonics
that is analogy of what is called a Gaunt coefficient for complex
spherical harmonics.  We will use the notation
$C(j_1\mu_1|j_1'\mu_1',\lambda \mu)$ for these  Gaunt coefficients 
here and in the two-electron integrals.
\begin{equation}
\begin{split}
& C(\ell m;\ell' m';\lambda \mu) = \\
& \int 
X_{\ell,m} (\mathbf{\hat{r}})
X_{\ell',m'} (\mathbf{\hat{r}})
X_{\lambda,\mu} (\mathbf{\hat{r}})
d\mathbf{\hat{r}} 
\end{split}
\end{equation}
Unlike the usual Gaunt coefficients, for which two of the spherical
harmonics are complex conjugated in the integral, these are symmetric
with respect to any permutation of pair of angular momentum indices.
With these we can assemble the basic integral
\begin{equation}
\boxed{
\begin{aligned}
& V^{nuc}_{i, \, \ell, \,m; \, j, \,\ell', \, m'}(\mathbf{R}_\beta) = \\
& \sum_{\lambda=|\ell-\ell'|}^{\ell+\ell'}
  U^\lambda_{i,j}(R_\beta) \\
& \times \left[
 \frac{4\pi} {2\lambda+1} 
\sum_{\mu=-\lambda}^{\lambda} X_{\lambda,\mu}(\mathbf{\hat{R}}_\beta)
C(\ell m;\ell' m';\lambda \mu)   \right] \\
\end{aligned}
}
\end{equation}
Note that the Gaunt coefficients are real.

That leaves us with the radial integral 
\begin{equation}
U^\lambda_{i,j}(R_\beta) = \int_0^{r_{max}} \chi_{i}(r) \chi_{j}(r) 
\frac{r_<^\lambda}{r_>^{\lambda+1}} 
\end{equation}
where the $r_<$ and $r_>$  refer to the less or greater of $r$ and $R_\beta$.
Using the Poisson equation approach, Eq.(66b) in the review \cite{TopicalReview2004}
(where there is a typo, the sum should be over $m$ not over $i$)
we get an explicit expression for this radial integral
\begin{equation}
\boxed{
\begin{aligned}
U^\lambda_{i,j}(R_\beta) & =  \\
\delta_{i,j}
& \left[
(2 \lambda +1) \sum_{m=1}^N 
\frac{\chi_m(R_\beta)}{R_\beta}
\left[ T^{(\lambda)}_{m,j}\right]^{-1} 
\frac{1}{r_{j}\sqrt{\rm{w}_{j}}} 
 \right. \\
& \left.  
 + \frac{R_\beta^{\lambda} r_{j}^\lambda }{r_{max}^{2\lambda+1}} 
\right]
\end{aligned}
}
\end{equation}
Note that the nuclear attraction integrals are diagonal in the 
DVR radial functions, but they are not diagonal in the angular momentum
indices.  

\subsubsection{Transformation of two-electron integrals}
With these formulas we can construct the matrix of the one-electron Hamiltonian,
$h_{i \ell m, j \ell' m'}$.  We could construct a nonspherical orbital basis by projecting 
out the doubly occupied orbitals in the target and then diagonalizing it to get orbitals of the 
form
\begin{equation}
\phi_a(\mathbf{r}) = \sum_{i,l,m} R^{(a)}_{ilm} \frac{\chi_i(r)}{r} X_{lm} (  \mathbf{\hat{r}} )
\end{equation} 
But this is a large matrix. Suppose we need to use up to $\ell = 10$, so that there would be $11^2 = 121$.  If there are 500 radial DVR points, then that matrix is 60,500 $\times$ 60,500, and we would need all of the eigenvectors.  That might be possible, but it doesn't seem optimum.   There are at least one other way to proceed to generate an orbital basis that is orthogonal to the doubly occupied orbitals.

Our projection operator to project out the doubly occupied orbitals from a one-electron function in Eq.(\ref{eq:Qonebody}) has the form
\begin{equation}
\begin{split}
\hat{Q}&=  \sum_{i ,\ell,m, j, \ell',m'} \frac{\chi_i(r)}{r} X_{\ell m} (  \mathbf{\hat{r}} )\hat{Q}_{i,\ell,m ; j, \ell',m'}   \frac{ \chi_j(r') }{r'} X_{\ell' m'} (  \mathbf{\hat{r'}} ) \\
\end{split}
\label{eq:Q1}
\end{equation}
It is notably \textbf{not} diagonal in angular momenta.  If we were to orthogonalize the basis in each angularm momentum block $\ell,m$ to the $\ell,m$ component of an orbital $\phi_a(\mathbf{r})$, the resulting basis would be orthogonal to $\phi_a(\mathbf{r})$ but also to other parts of the space not orthogonal to that orbital.  So we cannot project out one angular momentum component of $\hat{Q}$ and contruct the orthogonalized basis one partial wave at a time.  However we can contruct a basis that can be used to filter out high lying one-body energy eigenvectors, by starting with diagonalizing the diagonal blocks of the one-body Hamiltonian.

\begin{enumerate}
\item{ Diagonalize each diagonal block with respect to $\ell, m$ of the one-body Hamiltonian in the primitive DVR basis
\begin{equation}
\begin{split}
 h_{i,\ell,m, j, \ell',m'} &\equiv \int \frac{\chi_i(r)}{r} X_{\ell m}(\mathbf{\hat{r}}) \, \hat{h}(\mathbf{r}) \,  \frac{\chi_j(r)}{r} X_{\ell' m'}(\mathbf{\hat{r}}) d^3r\\
& \sum_{j}h_{i,\ell,m, j, \ell,m} R^{(n)}_{j,\ell m} = \epsilon_{n}^{\ell m} \, R^{(n)}_{i,\ell, m} \\
&\mathbf{R}_{\ell m} = \left(  \vec{R}^{(1)}_{\ell m},  \vec{R}^{(2)}_{\ell m},  \vec{R}^{(3)}_{\ell m}, \cdots, \vec{R}^{(N_{DVR})}_{\ell m} \right)
 \end{split}
\end{equation}
Each $\ell,m$ block of $h$ is of dimension $N_{DVR} \times N_{DVR}$ so for $\ell = 10$ this would be 121 $500 \times 500$ diagonalizations.
}
\item{Arrange the eigenvectors of the individual blocks to make a complete orbital space basis
\begin{equation}
\boldsymbol{\mathcal{R}}=\begin{pmatrix}
\mathbf{R}_{0 0} & 0&0&0& \cdots \\
0&\mathbf{R}_{1 -1} &0&0&\cdots \\
0&0&\mathbf{R}_{1 0} &0&\cdots \\
0&0 &0&\mathbf{R}_{1 0}&\cdots \\
\vdots&\vdots&\vdots&\vdots&\ddots
\end{pmatrix}
\end{equation}
The dimension of this matrix is $N_{DVR} N_{\ell m} \times N_{DVR} N_{\ell m}$ if all the eigenvectors are kept. If we want to drop eigenvectors 
with high values of $\epsilon_{n}^{\ell m}$, we drop the corresponding columns.  The matrix is indexed as
$\mathcal{R}_{i \ell m}^{a}$, with $a = 1,2 \cdots \sum_{\ell m}M_{\ell m}$, where $M_{\ell m}$ is the number of eigenvectors kept in  each $\ell,m$ block.
}
\item{Apply the projection operator
\begin{equation}
\tilde{\mathcal{R}}_{i,\ell,m}^a=\sum_{j,\ell',m'} \hat{Q}_{i,\ell,m ; j, \ell',m'} \mathcal{R}_{j,\ell',m'}^{a}
\end{equation}
This is the projected basis. Note that for any given value of $a$, $ \mathcal{R}_{j,\ell',m'}^{a}$ has nonzero values for only one value of $\ell' m'$, where as the projected functions given by $\tilde{\mathcal{R}}_{i,\ell,m}^a$ have nonzero values in all $\ell m$ blocks.
}
\end{enumerate}
This basis does \textbf{not} diagonalize the diagonal blocks of the one-body Hamiltonian any longer.  Any way of building an orthogonal orbital basis, other than diagonalizing the full, projected one-body Hamiltonian matrix and finding all the eigenvectors, will leave coupling in the one-body Hamiltonian matrix.  The original one-body Hamiltonian $h_{i \ell m;j \ell' m'}$ also has couplings between the $\ell m$ blocks because of the nonspherical nuclear potential and Coulomb and exchange operators from the frozen orbitals.


\subsubsection{$\left( 2 \hat{J}_o - \hat{K}_o \right)$}
This finishes the one electron integrals.


% appendices may preceed or follow the references
\appendix
\section{Identities involving angular momentum}
\label{sec:angmom}
Definition of real spherical harmonics in terms of the complex spherical harmonics with the Condon phases,
\begin{equation}
\begin{split}
&Y_{l,m}(\theta,\phi) =\\
&\begin{cases} (-1)^m  \left[ \frac{2l+1}{2} \frac{(l-|m|)!}{(l-|m|)!} \right]^{1/2} P_l^{|m|}(\cos \theta) \frac{1}{\sqrt{2\pi}} e^{i\, m \, \phi} & m>0 \\ 
\qquad \quad   \left[ \frac{2l+1}{4 \pi}  \right]^{1/2} P_l^{|m|}(\cos \theta)  & m=0 \\
\qquad \quad   \left[ \frac{2l+1}{2} \frac{(l-|m|)!}{(l-|m|)!} \right]^{1/2} P_l^{|m|}(\cos \theta)   \frac{1}{\sqrt{2\pi}} e^{i\, m \, \phi} & m < 0
\end{cases}
\end{split}
\end{equation}
is
\begin{equation}
\begin{split}
&X_{l,0}(\theta,\phi) = Y_{l,0}(\theta,\phi) \\
&X_{l,+|m|}(\theta,\phi) = \frac{1}{\sqrt{2}} \left((-1)^mY_{l,|m|}(\theta,\phi) + Y_{l,-|m|}(\theta,\phi) \right)\\
&X_{l,-|m|}(\theta,\phi) = \frac{1}{i\sqrt{2}} \left((-1)^mY_{l,|m|}(\theta,\phi) -Y_{l,-|m|}(\theta,\phi) \right)\\
& Y_{l,|m|}(\theta,\phi) = \frac{1}{\sqrt{2}} (-1)^m (X_{l,+|m|}(\theta,\phi)+i X_{l,-|m|}(\theta,\phi)) \\
& Y_{l,-|m|}(\theta,\phi) = \frac{1}{\sqrt{2}}  (X_{l,+|m|}(\theta,\phi)-iX_{l,-|m|}(\theta,\phi)) \\
\end{split}
\end{equation}
This definition makes the $+|m|$ real spherical harmonic the cosine-like one and the $-|m|$ real spherical harmonic the sine-like one.

The multipole expansion of $1/r_{12}$ is then the same in terms of complex or spherical harmonics 
\begin{equation}
\begin{split}
\frac{1}{| \mathbf{r}- \mathbf{r}'|} &= 
 \sum_{\lambda,\mu} \frac{4 \pi}{2\lambda +1}
Y_{\lambda,\mu}(\mathbf{\hat{r}})^*
Y_{\lambda,\mu}(\mathbf{\hat{r}}')
\frac{r^\lambda}{r^{\lambda+1}} \\
&= 
 \sum_{\lambda,\mu} \frac{4 \pi}{2\lambda +1}
X_{\lambda,\mu}(\mathbf{\hat{r}})
X_{\lambda,\mu}(\mathbf{\hat{r}}')
\frac{r^\lambda}{r^{\lambda+1}} 
\end{split}
\end{equation}
as can be seen by substitution.

Using
\begin{equation}
Y^*_{\ell,m} = (-1)^m Y_{\ell,-m} 
\end{equation}
The real spherical harmonics can also be written
\begin{equation}
\begin{split}
&X_{l,0}(\theta,\phi) = Y_{l,0}(\theta,\phi) \\
&X_{l,+|m|}(\theta,\phi) = \frac{(-1)^m}{\sqrt{2}} \left(Y_{l,|m|}(\theta,\phi) + Y^*_{l,|m|}(\theta,\phi) \right)\\
&X_{l,-|m|}(\theta,\phi) = \frac{(-1)^m}{i\sqrt{2}} \left(Y_{l,|m|}(\theta,\phi) -Y^*_{l,|m|}(\theta,\phi) \right)\\
& Y_{l,|m|}(\theta,\phi) = \frac{(-1)^m }{\sqrt{2}} (X_{l,+|m|}(\theta,\phi)+i X_{l,-|m|}(\theta,\phi)) \\
& Y^*_{l,|m|}(\theta,\phi) = \frac{(-1)^m}{\sqrt{2}}  (X_{l,+|m|}(\theta,\phi)-iX_{l,-|m|}(\theta,\phi)) \\
\end{split}
\end{equation}

We will use the notation
$C(j_1\mu_1|j_1'\mu_1',\lambda \mu)$ for the Gaunt coefficients 
in terms of real spherical harmonics
\begin{equation}
\begin{split}
& C(j_1\mu_1|j_1'\mu_1',\lambda \mu) = \\
& \int 
X_{\lambda,\mu} (\mathbf{\hat{r}})
X_{j_1',\mu_1'} (\mathbf{\hat{r}}) X_{j_1,\mu_1} (\mathbf{\hat{r}})
d\mathbf{\hat{r}}  
\end{split}
\end{equation}
These integrals can be related to the basic integral of three complex spherical
harmonics, but I haven't found a simpler expression than one involving all eight terms
for each case of positive or negative m values.
\begin{equation}
\begin{split}
 \int 
Y_{\lambda,\mu} (\mathbf{\hat{r}})&
Y_{j_1',\mu_1'} (\mathbf{\hat{r}})
Y_{j_1,\mu_1} (\mathbf{\hat{r}})
d\mathbf{\hat{r}}  = \\
&
\left[\frac{(2j_1' + 1)(2j_1+1)(2\lambda+1)}{4\pi}\right]^{1/2} \\
& \times 
\begin{pmatrix}
\lambda&j_1'&j_1\\
0&0&0
\end{pmatrix}
\begin{pmatrix}
\lambda&j_1'&j_1\\
-\mu&-\mu_1'&\mu_1
\end{pmatrix}
\end{split}
\end{equation}


\section{Other identies}
% make bibliography
\bibliography{DVR_doublePI.bib}
%
\end{document}  % 


